---
title: "Log-normal regression with gamlss"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Log-normal regression with gamlss}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



### Loading data

To facilitate the analysis we have loaded the databases into the **BlauCat** package as an internal dataset labelled **blaucat_data**. That is, the data will be available to the user once the package is loaded with the **library** or **require** functions.

To fill in some of the missing organic carbon data in the original dataset we used the transformation $CO = 0.45\times MO$, where $CO$ indicates organic carbon and $MO$ stands for organic matter. This way, we ended up with 56 valid points.

```{r setup}
library(BlauCat)
```


### Log-normal regression

After some trials, we have turned to a log-normal regression to account for the long positive tail of the data. We used the **gamlsss* R package for this, picking the log-normal family distribution with 

As a first step we will select the predictors that will be used in the regression. Factors are defined as such.

```{r select}
df <- blaucat_dat[, c("Carboni orgànic del sediment_mean",
                      "Codi CORINE 2024",
                      "Cota per dalt",
                      "Percentatge de sorra_mean",
                      "Temperatura mitjana de l'aire",
                      "Precipitació acumulada",
                      "Elevació respecte al nivell del mar",
                      "Distància a zones agrícoles",
                      "Distància a zones urbanes")]
df$`Cota per dalt` <- as.numeric(df$`Cota per dalt`)

# Get distance to nearest water body.
blaucat_dat$`Distància a badia` <- as.numeric(blaucat_dat$`Distància a badia`)
df$`Distància aigua` <- with(blaucat_dat,
                             pmin(`Distància al mar`, `Distància a badia`,
                                  `Distància al riu`, na.rm = TRUE))

# Recodify CORINE.
df <- df |>
  dplyr::mutate(`Codi CORINE 2024` = dplyr::recode(`Codi CORINE 2024`,
    "Canyissars i formacions anàlogues" = "Canyissars",
    "Formacions de Cladium mariscus" = "Cladium mariscus",
    "Vegetació (salicornars. prats. jonqueres...) dels sòls salins. poc o molt humits o. fins i tot. inundats. del litoral" = "Prats i jonqueres"
  ))

df <- na.omit(df)
```

<br>

We have a look at relationships between variables.

```{r mapa-general}
# General view.
blaucat_dat$`UTM X` <- as.numeric(blaucat_dat$`UTM X`)
blaucat_dat$`UTM Y` <- as.numeric(blaucat_dat$`UTM Y`)
xy <- terra::vect(blaucat_dat, geom = c("UTM X", "UTM Y"), crs = "EPSG:25831")
terra::values(xy) <- NULL

spain_adm1 <- geodata::gadm(country = "ESP", level = 1, path = tempdir())
catalonia <- spain_adm1[spain_adm1$NAME_1 == "Cataluña", ] |>
  terra::project("EPSG:25831")

library(terra)
library(mapSpain)
library(geodata)

# 1. Fetch Catalonia boundary and major rivers
cat_sf <- esp_get_ccaa(ccaa = "Catalunya")
rivers_sf <- esp_get_rivers()

cat_sv <- vect(cat_sf)
rivers_sv <- vect(rivers_sf)

# 2. Fetch SRTM / Shuttle elevation data for Spain (returns a SpatRaster)
dem_spain <- elevation_30s(country = "ESP", path = tempdir())

# 3. Project all layers to UTM 31N (EPSG:25831)
target_crs <- "EPSG:25831"

cat_utm <- project(cat_sv, target_crs)
rivers_utm <- project(rivers_sv, target_crs)

# Project DEM raster using bilinear interpolation for continuous values
dem_utm <- project(dem_spain, target_crs, method = "bilinear")

# 4. Crop and mask DEM & rivers to Catalonia's boundary
dem_cat <- crop(dem_utm, cat_utm) |> mask(cat_utm)
rivers_cat <- crop(rivers_utm, cat_utm)

# 5. Plot topographic map with river overlay
# Plot raster elevation base layer
terra::plot(
  dem_cat,
  col = terrain.colors(50),
  main = "MDE (m)",
  pax = list(cex.axis = 1.5),
  plg = list(cex = 1.5)
)

# Overlay border vector
plot(cat_utm, border = "gray20", lwd = 1, add = TRUE)

# Overlay river vector lines
plot(rivers_cat, col = "dodgerblue3", lwd = 2, add = TRUE)

# terra::plot(catalonia)
terra::plot(xy, add = TRUE, cex = 1.5)

xy_wgs84 <- terra::project(xy, "EPSG:4326")

writeVector(xy_wgs84, "../inst/google/catalonia_data.kml", filetype = "KML", overwrite = TRUE)

```



Then, we log-transform the distance predictors.

```{r log-transform}
df$`Distància aigua` <- log(df$`Distància aigua`) # There are a few zeros.
df$`Distància a zones agrícoles` <- log(df$`Distància a zones agrícoles`)
df$`Distància a zones urbanes` <- log(df$`Distància a zones urbanes`)
```

<br>

 We will also rename the variables from Catalan names to English ones.
 
```{r rename}
names(df) <- c("Carbon", "Corine", "Upperlimit", "MeanSand", "Temp", "Prec",
               "Elev", "DistAgric", "DistUrb", "DistWater")
df$Corine <- factor(df$Corine)
```


We plot 
```{r boxplot}
boxplot(df$Carbon ~ df$Corine, xlab = "CORINE", ylab = "Carbon")

library(ggplot2)

ggplot(df, aes(x = factor(Corine), y = Carbon)) +
  geom_violin(fill = "skyblue", color = "dodgerblue4", alpha = 0.7, trim = FALSE) +
  labs(
    x = "CORINE Category",
    y = "Carbon",
    title = "Distribution of Carbon by CORINE Land Cover"
  ) +
  theme_minimal()
```

<br>

Finally, we try a regression model with the **gamlss** package, using a log-normal distribution for the residuals. The dispersion variable $\sigma$ is modelled simply by a constant.

```{r model}
# Regression package.
suppressWarnings(suppressPackageStartupMessages(library(gamlss,
                                                        warn.conflicts = FALSE,
                                                        verbose = FALSE,
                                                        quietly = TRUE)))
# Rescale package.
suppressWarnings(suppressPackageStartupMessages(library(arm,
                                                        warn.conflicts = FALSE,
                                                        verbose = FALSE,
                                                        quietly = TRUE)))
library(broom.mixed)

# Do not rescale categorical variables.
model <- lm.beta::lm.beta(lm(formula = log(Carbon) ~ ., data = df))
tidy_model <- tidy(model)
knitr::kable(tidy_model, digits = 4, caption = "GAMLSS Model Results")
```

<br>

As an additional performance indicator we can calculate the linear correlation between prediction and response, noting that this may be unreliable due to the presence of heterocedasticity.

```{r linearcor}
library(kableExtra)
pr <- exp(predict(model) + 0.5*sd(resid(model))^2)
knitr::kable(data.frame(Pearson = cor(df$Carbon, pr),
  Spearman = cor(df$Carbon, pr, method = "spearman")), digits = 4,
  caption = "Predicted vs. response") |> kable_styling(full_width = TRUE)
```



Now, with random forest.

```{r ranger}
library(ranger)
library(knitr)
library(dplyr)
library(pdp)
library(ggplot2)
library(gridExtra)

# 1. Fit ranger model with importance enabled
rf_model <- ranger(
  formula = log(Carbon) ~ .,
  data = df,
  num_trees = 1000,
  importance = "permutation"
)

# 2. Extract importance values into a data frame
importance_df <- data.frame(
  Variable = names(ranger::importance(rf_model)),
  Importance = ranger::importance(rf_model)
) |> 
  arrange(desc(Importance))

# 3. Format as a table
kable(
  importance_df, 
  digits = 4, 
  row.names = FALSE,
  caption = "Ranger Random Forest Variable Importance"
)

# -------------------------------------------------------------
# 4. Define Bootstrapped PDP Function
# -------------------------------------------------------------

get_bootstrapped_pdp <- function(model, var, data, n_boot = 50, grid_resolution = 20) {
  
  if (is.numeric(data[[var]])) {
    grid_vals <- seq(min(data[[var]], na.rm = TRUE), max(data[[var]], na.rm = TRUE), length.out = grid_resolution)
  } else {
    grid_vals <- unique(data[[var]])
  }
  
  boot_results <- replicate(n_boot, {
    boot_df <- data[sample(nrow(data), replace = TRUE), ]
    pd <- partial(model, pred.var = var, train = boot_df, grid.resolution = grid_resolution)
    pd$yhat
  })
  
  data.frame(
    x = grid_vals,
    yhat = rowMeans(boot_results),
    lower = apply(boot_results, 1, quantile, probs = 0.025),
    upper = apply(boot_results, 1, quantile, probs = 0.975)
  )
}

# -------------------------------------------------------------
# 5. Generate PDP Plots with Confidence Bands & Slanted Labels
# -------------------------------------------------------------

predictors <- c("Corine", "Upperlimit", "MeanSand", "Temp", "Prec", 
                "Elev", "DistAgric", "DistUrb", "DistWater")

set.seed(123)

pdp_plots <- lapply(predictors, function(var) {
  
  pdp_data <- get_bootstrapped_pdp(rf_model, var = var, data = df, n_boot = 50)
  
  # Check if variable is categorical/factor or explicitly "Corine"
  is_categorical <- is.factor(df[[var]]) || is.character(df[[var]]) || var == "Corine"
  
  if (!is_categorical) {
    # Numeric Plot: Ribbon + Line
    p <- ggplot(pdp_data, aes(x = x, y = yhat)) +
      geom_ribbon(aes(ymin = lower, ymax = upper), fill = "skyblue", alpha = 0.4) +
      geom_line(color = "dodgerblue4", linewidth = 1) +
      theme_minimal()
  } else {
    # Categorical Plot: PointRange with Slanted X Labels
    p <- ggplot(pdp_data, aes(x = factor(x), y = yhat)) +
      geom_pointrange(aes(ymin = lower, ymax = upper), color = "dodgerblue4", size = 0.6) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(
          angle = 45,     # Slanted angle
          vjust = 1,      # Vertical alignment to clear the axis
          hjust = 1,      # Horizontal alignment to anchor text right-side
          size = 8        # Slightly smaller size for longer level names
        )
      )
  }
  
  p + labs(
    title = var,
    x = var,
    y = "log(Carbon)"
  ) +
  theme(
    plot.title = element_text(size = 11, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 9)
  )
})

# 6. Display PDP plots in a 3x3 grid
grid.arrange(grobs = pdp_plots, ncol = 3)
```
 
