---
title: "Log-normal gamlss and random forest models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Log-normal regression with gamlss}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



### Loading data

To facilitate the analysis we have loaded the databases into the **BlauCat** package as an internal dataset labelled **blaucat_data**. That is, the data will be available to the user once the package is loaded with the **library** or **require** functions.

To fill in some of the missing organic carbon data in the original dataset we used the transformation $CO = 0.45\times MO$, where $CO$ indicates organic carbon and $MO$ stands for organic matter. This way, we ended up with 56 valid points.

```{r setup}
library(BlauCat)
```


### Data pre-processing

As a first step we will select the predictors that will be used in the regression. Factors are defined as such.

```{r select}
df <- blaucat_dat[, c("Carboni orgànic del sediment_mean",
                      "Codi CORINE 2024",
                      "Cota per dalt",
                      "Percentatge de sorra_mean",
                      "Temperatura mitjana de l'aire",
                      "Precipitació acumulada",
                      "Elevació respecte al nivell del mar",
                      "Distància a zones agrícoles",
                      "Distància a zones urbanes")]
df$`Cota per dalt` <- as.numeric(df$`Cota per dalt`)

# Get distance to nearest water body.
blaucat_dat$`Distància a badia` <- as.numeric(blaucat_dat$`Distància a badia`)
df$`Distància aigua` <- with(blaucat_dat,
                             pmin(`Distància al mar`, `Distància a badia`,
                                  `Distància al riu`, na.rm = TRUE))

# Recodify CORINE.
df <- df |>
  dplyr::mutate(`Codi CORINE 2024` = dplyr::recode(`Codi CORINE 2024`,
    "Canyissars i formacions anàlogues" = "Canyissars",
    "Formacions de Cladium mariscus" = "Cladium mariscus",
    "Vegetació (salicornars. prats. jonqueres...) dels sòls salins. poc o molt humits o. fins i tot. inundats. del litoral" = "Prats i jonqueres"
  ))

df <- na.omit(df)
```

<br>

We have a look at relationships between variables.

```{r mapa-general}
# General view.
blaucat_dat$`UTM X` <- as.numeric(blaucat_dat$`UTM X`)
blaucat_dat$`UTM Y` <- as.numeric(blaucat_dat$`UTM Y`)
xy <- terra::vect(blaucat_dat, geom = c("UTM X", "UTM Y"), crs = "EPSG:25831")
terra::values(xy) <- NULL

spain_adm1 <- geodata::gadm(country = "ESP", level = 1, path = tempdir())
catalonia <- spain_adm1[spain_adm1$NAME_1 == "Cataluña", ] |>
  terra::project("EPSG:25831")

library(terra)
library(mapSpain)
library(geodata)

# 1. Fetch Catalonia boundary and major rivers
cat_sf <- esp_get_ccaa(ccaa = "Catalunya")
rivers_sf <- esp_get_rivers()

cat_sv <- vect(cat_sf)
rivers_sv <- vect(rivers_sf)

# 2. Fetch SRTM / Shuttle elevation data for Spain (returns a SpatRaster)
dem_spain <- elevation_30s(country = "ESP", path = tempdir())

# 3. Project all layers to UTM 31N (EPSG:25831)
target_crs <- "EPSG:25831"

cat_utm <- project(cat_sv, target_crs)
rivers_utm <- project(rivers_sv, target_crs)

# Project DEM raster using bilinear interpolation for continuous values
dem_utm <- project(dem_spain, target_crs, method = "bilinear")

# 4. Crop and mask DEM & rivers to Catalonia's boundary
dem_cat <- crop(dem_utm, cat_utm) |> mask(cat_utm)
rivers_cat <- crop(rivers_utm, cat_utm)

# 5. Plot topographic map with river overlay
# Plot raster elevation base layer
terra::plot(
  dem_cat,
  col = terrain.colors(50),
  main = "MDE (m)",
  pax = list(cex.axis = 1.5),
  plg = list(cex = 1.5)
)

# Overlay border vector
plot(cat_utm, border = "gray20", lwd = 1, add = TRUE)

# Overlay river vector lines
plot(rivers_cat, col = "dodgerblue3", lwd = 2, add = TRUE)

# terra::plot(catalonia)
terra::plot(xy, add = TRUE, cex = 1.5)

xy_wgs84 <- terra::project(xy, "EPSG:4326")

writeVector(xy_wgs84, "../inst/google/catalonia_data.kml", filetype = "KML", overwrite = TRUE)

```

<br>

Then, we log-transform the distance predictors.

```{r log-transform}
df$`Distància aigua` <- log(df$`Distància aigua`) # There are a few zeros.
df$`Distància a zones agrícoles` <- log(df$`Distància a zones agrícoles`)
df$`Distància a zones urbanes` <- log(df$`Distància a zones urbanes`)
```

<br>

 We will also rename the variables from Catalan names to English ones.
 
```{r rename}
names(df) <- c("Carbon", "Corine", "Upperlimit", "MeanSand", "Temp", "Prec",
               "Elev", "DistAgric", "DistUrb", "DistWater")
df$Corine <- factor(df$Corine)
```


We plot dependent variable vs. categorical (CORINE).

```{r violin-plot}
library(ggplot2)

ggplot(df, aes(x = factor(Corine), y = Carbon, fill = factor(Corine))) +
  # 1. Violin shapes with custom transparency and outline
  geom_violin(color = "grey30", alpha = 0.7, trim = FALSE, show.legend = FALSE) +
  
  # 2. Narrow inner boxplot to display median and quartiles
  geom_boxplot(width = 0.15, fill = "white", color = "grey20", outlier.shape = NA, alpha = 0.9) +
  
  # 3. Jittered observations to show actual sample distribution
  geom_jitter(width = 0.08, alpha = 0.3, color = "black", size = 1.2) +
  
  # 4. Color palette per category
  scale_fill_brewer(palette = "Set2") +
  
  # 5. Clean, publication-ready formatting
  labs(
    x = "CORINE Category",
    y = "Carbon Content",
    title = "Distribucio de Carbó vs. CORINE",
    subtitle = "Violins indiquen densitat; caixes internes indiquen mediana i IQR"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 14, margin = margin(b = 4)),
    plot.subtitle = element_text(color = "grey40", size = 10, margin = margin(b = 10)),
    axis.title = element_text(face = "bold", size = 11),
    axis.text = element_text(size = 10, color = "black"),
    panel.grid.major.x = element_blank(), # Remove vertical gridlines
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "black", linewidth = 0.8)
  )

```

Plot the numeric variables.

```{r scatter-plots}
library(tidyr)

# Pivot continuous variables (excluding Carbon and Corine)
df_long <- pivot_longer(
  df, 
  cols = -c(Carbon, Corine), 
  names_to = "Variable", 
  values_to = "Value"
)

ggplot(df_long, aes(x = Value, y = Carbon)) +
  geom_point(alpha = 0.7, color = "#2c3e50") +
  facet_wrap(~ Variable, scales = "free_x") +
  theme_bw() + # Base theme with full box borders
  theme(
    # Panel box formatting
    panel.background = element_rect(fill = "#f8f9fa"),
    panel.border = element_rect(color = "#333333", fill = NA, linewidth = 1),
    panel.spacing = unit(1.2, "lines"), # Increase physical gap between panels
    
    # Strip (header) formatting
    strip.background = element_rect(fill = "#e2e8f0", color = "#333333", linewidth = 1),
    strip.text = element_bold_text <- element_text(size = 10, face = "bold", color = "#1a202c"),
    
    # Grid lines
    panel.grid.major = element_line(color = "#e2e8f0", linewidth = 0.5),
    panel.grid.minor = element_blank()
  ) +
  labs(
    x = "Predictor Value", 
    y = "Carbon", 
    title = "Carbon vs Continuous Predictors"
  )
```



<br>

### Linear regression with log-transformation of dependent variable.

We try a regression model assuming a log-normal distribution for the residuals.

```{r model-lm}
lm_model <- lm.beta::lm.beta(lm(formula = log(Carbon) ~ ., data = df))
tidy_model <- broom::tidy(lm_model)
knitr::kable(tidy_model, digits = 4, caption = "Model de regressió")
```

<br>

Plot diagnostic for linear regression model.

```{r lm-diagnostics}

library(ggfortify)

autoplot(lm_model, 
         which = 1:4, 
         ncol = 2, 
         colour = "#2c3e50", 
         smooth.colour = "#e74c3c", 
         smooth.linewidth = 0.8) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    panel.border = element_rect(color = "black", linewidth = 0.8),
    panel.grid.minor = element_blank(),
    panel.spacing = unit(1, "lines")
  )
```

<br>

As an additional performance indicator we can calculate the linear correlation between prediction and response, noting that this may be unreliable due to the presence of heterocedasticity.

```{r linearcor}
library(kableExtra)
pr <- exp(predict(lm_model) + 0.5*sd(resid(lm_model))^2)
knitr::kable(data.frame(Pearson = cor(df$Carbon, pr),
  Spearman = cor(df$Carbon, pr, method = "spearman")), digits = 4,
  caption = "Predicted vs. response") |> kable_styling(full_width = TRUE)
```



### Random forest model (ranger package)


```{r model-ranger}
library(ranger)
library(knitr)
library(dplyr)
library(pdp)
library(gridExtra)

# 1. Fit ranger model with importance enabled
rf_model <- ranger(
  formula = log(Carbon) ~ .,
  data = df,
  num_trees = 1000,
  importance = "permutation"
)

# 2. Extract importance values into a data frame
importance_df <- data.frame(
  Variable = names(ranger::importance(rf_model)),
  Importance = ranger::importance(rf_model)
) |> 
  arrange(desc(Importance))

# 3. Format as a table
kable(
  importance_df, 
  digits = 4, 
  row.names = FALSE,
  caption = "Ranger Random Forest Variable Importance"
)

```



```{r PDP-plots}

# Define Bootstrapped PDP Function

get_bootstrapped_pdp <- function(model, var, data, n_boot = 50, grid_resolution = 20) {
  
  if (is.numeric(data[[var]])) {
    grid_vals <- seq(min(data[[var]], na.rm = TRUE), max(data[[var]], na.rm = TRUE), length.out = grid_resolution)
  } else {
    grid_vals <- unique(data[[var]])
  }
  
  boot_results <- replicate(n_boot, {
    boot_df <- data[sample(nrow(data), replace = TRUE), ]
    pd <- partial(model, pred.var = var, train = boot_df, grid.resolution = grid_resolution)
    pd$yhat
  })
  
  data.frame(
    x = grid_vals,
    yhat = rowMeans(boot_results),
    lower = apply(boot_results, 1, quantile, probs = 0.025),
    upper = apply(boot_results, 1, quantile, probs = 0.975)
  )
}

# Generate PDP Plots with Confidence Bands & Slanted Labels

predictors <- c("Corine", "Upperlimit", "MeanSand", "Temp", "Prec", 
                "Elev", "DistAgric", "DistUrb", "DistWater")

set.seed(123)

pdp_plots <- lapply(predictors, function(var) {
  
  pdp_data <- get_bootstrapped_pdp(rf_model, var = var, data = df, n_boot = 50)
  
  # Check if variable is categorical/factor or explicitly "Corine"
  is_categorical <- is.factor(df[[var]]) || is.character(df[[var]]) || var == "Corine"
  
  if (!is_categorical) {
    # Numeric Plot: Ribbon + Line
    p <- ggplot(pdp_data, aes(x = x, y = yhat)) +
      geom_ribbon(aes(ymin = lower, ymax = upper), fill = "skyblue", alpha = 0.4) +
      geom_line(color = "dodgerblue4", linewidth = 1) +
      theme_minimal()
  } else {
    # Categorical Plot: PointRange with Slanted X Labels
    p <- ggplot(pdp_data, aes(x = factor(x), y = yhat)) +
      geom_pointrange(aes(ymin = lower, ymax = upper), color = "dodgerblue4", size = 0.6) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(
          angle = 45,     # Slanted angle
          vjust = 1,      # Vertical alignment to clear the axis
          hjust = 1,      # Horizontal alignment to anchor text right-side
          size = 8        # Slightly smaller size for longer level names
        )
      )
  }
  
  p + labs(
    title = var,
    x = var,
    y = "log(Carbon)"
  ) +
  theme(
    plot.title = element_text(size = 11, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 9)
  )
})

# Display PDP plots in a 3x3 grid
grid.arrange(grobs = pdp_plots, ncol = 3)
```


### Training and test.

```{r train-test}
nsimu <- 1000
p <- .9
carbon_rmse <- matrix(0, nsimu, 2)
for (i in 1:nsimu) {
  index <- sample_stratified(df, p = p, "Corine")
  
  df_train <- df[index, ]
  train_lm <- lm(log(Carbon) ~ ., data = df_train)
  train_rf <- ranger::ranger(log(Carbon) ~ ., num.trees = 1000, data =  df_train)
  
  df_test <- df[-index, ]
  lm_sd <- mean(resid(train_lm)^2)
  carbon_rmse[i, 1] <- sqrt(mean((exp(predict(train_lm, df_test)+.5*lm_sd^2) - df_test$Carbon)^2))
  carbon_rmse[i, 2] <- sqrt(mean((exp(predict(train_rf, data = df_test)$predictions+.5*lm_sd^2) - df_test$Carbon)^2))
}

# Check RMSE for training/test approach.
mean_rmse <- setNames(apply(carbon_rmse, 2,mean), c("Linear regression", "Random Forest"))
knitr::kable(mean_rmse)

```


