---
title: "Log-normal regression with gamlss"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Log-normal regression with gamlss}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



### Loading data

To facilitate the analysis we have loaded the databases into the **BlauCat** package as an internal dataset labelled **blaucat_data**. That is, the data will be available to the user once the package is loaded with the **library** or **require** functions.

To fill in some of the missing organic carbon data in the original dataset we used the transformation $CO = 0.45\times MO$, where $CO$ indicates organic carbon and $MO$ stands for organic matter. This way, we ended up with 56 valid points.

```{r setup}
library(BlauCat)
```


### Log-normal regression

After some trials, we have turned to a log-normal regression to account for the long positive tail of the data. We used the **gamlsss* R package for this, picking the log-normal family distribution with 

As a first step we will select the predictors that will be used in the regression. Factors are defined as such.

```{r select}
df <- blaucat_dat[, c("Carboni orgànic del sediment_mean",
                      "Ús del sòl del lloc de mostreig",
                      "Distància al mar",
                      "Distància a zones agrícoles",
                      "Distància a zones urbanes",
                      "Temperatura mitjana de l'aire",
                      "Precipitació acumulada",
                      "Elevació respecte al nivell del mar",
                      "Tipus de sòl")]
df$`Ús del sòl del lloc de mostreig` <- factor(df$`Ús del sòl del lloc de mostreig`)
df$`Tipus de sòl` <- factor(df$`Tipus de sòl`)
df <- na.omit(df)
```

<br>

Then, we log-transform the distance predictors.

```{r log-transform}
df$`Distància al mar` <- log(df$`Distància al mar`)
df$`Distància a zones agrícoles` <- log(df$`Distància a zones agrícoles`)
df$`Distància a zones urbanes` <- log(df$`Distància a zones urbanes`)
```

<br>

 We will also rename the variables from Catalan names to English ones. It turns out that one of the levels in factor "Soil" is empty, so we will refactor it.
 
```{r rename}
names(df) <- c("Carbon", "LandUs", "DistSea", "DistAgri", "DistUrba", "Temp", "Prec", "Elev", "Soil")
df$Soil <- factor(df$Soil)
```

<br>

Finally, we try a regression model with the **gamlss** package, using a log-normal distribution for the residuals. The dispersion variable $\sigma$ is modelled simply by a constant.

```{r model}
# Regression package.
suppressWarnings(suppressPackageStartupMessages(library(gamlss,
                                                        warn.conflicts = FALSE,
                                                        verbose = FALSE,
                                                        quietly = TRUE)))
# Rescale package.
suppressWarnings(suppressPackageStartupMessages(library(arm,
                                                        warn.conflicts = FALSE,
                                                        verbose = FALSE,
                                                        quietly = TRUE)))
df2 <- lapply(df, rescale)
df2$Carbon <- df$Carbon
m1 <- gamlss(formula = Carbon ~ ., sigma.formula = ~ 1,
             data = df2, family = LOGNO, control = gamlss.control(trace = FALSE))
library(broom.mixed)
tidy_m1 <- tidy(m1)
knitr::kable(tidy_m1, digits = 4, caption = "GAMLSS Model Results")
```

<br>

As an additional performance indicator we can calculate the linear correlation between prediction and response, noting that this may be unreliable due to the presence of heterocedasticity.

```{r linearcor}
library(kableExtra)
knitr::kable(data.frame(Pearson = cor(df$Carbon, predict(m1)),
  Spearman = cor(df$Carbon, predict(m1), method = "spearman")), digits = 4,
  caption = "Predicted vs. response") |> kable_styling(full_width = TRUE)
```
 
 
