---
title: "Log-normal regression with gamlss"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Log-normal regression with gamlss}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



### Loading data

To facilitate the analysis we have loaded the databases into the **BlauCat** package as an internal dataset labelled **blaucat_data**. That is, the data will be available to the user once the package is loaded with the **library** or **require** functions.

To fill in some of the missing organic carbon data in the original dataset we used the transformation $CO = 0.45\times MO$, where $CO$ indicates organic carbon and $MO$ stands for organic matter. This way, we ended up with 56 valid points.

```{r setup}
library(BlauCat)
```


### Log-normal regression

After some trials, we have turned to a log-normal regression to account for the long positive tail of the data. We used the **gamlsss* R package for this, picking the log-normal family distribution with 

As a first step we will select the predictors that will be used in the regression. Factors are defined as such.

```{r select}
df <- blaucat_dat[, c("Carboni orgànic del sediment_mean",
                      "Codi CORINE 2024",
                      "Cota per dalt",
                      "Percentatge de sorra_mean",
                      "Temperatura mitjana de l'aire",
                      "Precipitació acumulada",
                      "Elevació respecte al nivell del mar",
                      "Distància a zones agrícoles",
                      "Distància a zones urbanes")]
df$`Cota per dalt` <- as.numeric(df$`Cota per dalt`)

# Get distance to nearest water body.
blaucat_dat$`Distància a badia` <- as.numeric(blaucat_dat$`Distància a badia`)
df$`Distància aigua` <- with(blaucat_dat,
                             pmin(`Distància al mar`, `Distància a badia`,
                                  `Distància al riu`, `Distància a llacunes`, na.rm = TRUE))

# Recodify CORINE.
df <- df |>
  dplyr::mutate(`Codi CORINE 2024` = dplyr::recode(`Codi CORINE 2024`,
    "Canyissars i formacions anàlogues" = "Canyissars",
    "Formacions de Cladium mariscus" = "Cladium mariscus",
    "Vegetació (salicornars. prats. jonqueres...) dels sòls salins. poc o molt humits o. fins i tot. inundats. del litoral" = "Prats i jonqueres"
  ))

df <- na.omit(df)
```

<br>

Then, we log-transform the distance predictors.

```{r log-transform}
df$`Distància aigua` <- sqrt(df$`Distància aigua`) # There are a few zeros.
df$`Distància a zones agrícoles` <- sqrt(df$`Distància a zones agrícoles`)
df$`Distància a zones urbanes` <- sqrt(df$`Distància a zones urbanes`)
```

<br>

 We will also rename the variables from Catalan names to English ones.
 
```{r rename}
names(df) <- c("Carbon", "Corine", "Upperlimit", "MeanSand", "Temp", "Prec",
               "Elev", "DistAgric", "DistUrb", "DistWater")
```

<br>

Finally, we try a regression model with the **gamlss** package, using a log-normal distribution for the residuals. The dispersion variable $\sigma$ is modelled simply by a constant.

```{r model}
# Regression package.
suppressWarnings(suppressPackageStartupMessages(library(gamlss,
                                                        warn.conflicts = FALSE,
                                                        verbose = FALSE,
                                                        quietly = TRUE)))
# Rescale package.
suppressWarnings(suppressPackageStartupMessages(library(arm,
                                                        warn.conflicts = FALSE,
                                                        verbose = FALSE,
                                                        quietly = TRUE)))
library(broom.mixed)

# Do not rescale categorical variables.
i <- setdiff(colnames(df), c("Carbon", "Corine"))
df[i] <- lapply(df[i], rescale)
model <- lm(formula = log(Carbon) ~ ., data = df)
tidy_model <- tidy(model)
knitr::kable(tidy_model, digits = 4, caption = "GAMLSS Model Results")
```

<br>

As an additional performance indicator we can calculate the linear correlation between prediction and response, noting that this may be unreliable due to the presence of heterocedasticity.

```{r linearcor}
library(kableExtra)
pr <- exp(predict(model) + 0.5*sd(resid(model))^2)
knitr::kable(data.frame(Pearson = cor(df$Carbon, pr),
  Spearman = cor(df$Carbon, pr, method = "spearman")), digits = 4,
  caption = "Predicted vs. response") |> kable_styling(full_width = TRUE)
```

